Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-04-27 paper.tex
Update LS section

The ``succinct tree sequence'' is an ARG data structure
that is both general (in terms of the types of ancestry that can
be described) and computationally efficient~\citep{Wong2023-efficient}.
Originally developed to facilitate large-scale coalescent
simulations~\citep{Kelleher2016-wk}, the methods have been
extended and applied to forwards-time
simulations~\citep{Kelleher2018-xc,Haller2018-tree},
calculation of population genetics statistics~\citep{Ralph2020-efficiently}
and ARG inference~\citep{Kelleher2019-ba,Wohns2022-th}.
The succinct tree sequence is based on a simple tabular representation,
which defines a set of nodes, edges, sites and mutations. A \emph{node}
represents a particular genome, which may be an observed sample
or an inferred genetic ancestor. The genetic inheritance between
a pair of nodes along a segment of genome is defined by
the \emph{edge} $(\ell, r, p, c)$, which states that
child node $c$ inherited its genome from parent node $p$
from left coordinate $\ell$ to right coordinate $r$. A \emph{site}
defines the genomic position at which some variation occurs,
and a \emph{mutation} defines the site, node and derived state
that generates variation.
Given these tables, we can then efficiently construct the local
genealogical trees along the genome (arising from recombination),
and also perform a range of calculations efficiently by
reasoning about the differences between these local
trees~\citep{Kelleher2016-wk,Ralph2020-efficiently}. These
algorithms have led to peformance increases of several orders
of magnitude over the state-of-the-art in a range of
applications~\citep{Kelleher2016-wk,Kelleher2018-xc,Kelleher2019-ba,
Ralph2020-efficiently,Baumdicker2022-ep}.
The succinct tree sequence encoding
is also very concise, allowing, for example, for millions of
complete human genomes to be stored in a few gigabytes of
space~\citep{Kelleher2019-ba}.

@@

[JK: TODO, expand this next paragraph to two, explaining a bit more that
we're searching among every node in the ARG, using the efficient tree
implementation in tsinfer, and a few words about how we can use the
HMM more, to do things like stochastic traceback, etc.]

Recombination is inferred using an efficient HMM implementation of the
LS model. Copying paths are estimated from SARS-CoV-2 genomes sampled over
time, with younger genomes copying (i.e. inheriting segments and mutations)
from older genomes. Here, we were able to apply the LS model on millions of
SARS-CoV-2 genomes because of a highly efficient, exact HMM implementation of
the LS model that is a core component of the tsinfer software. For this
application, we have parameterised the HMM using a mismatch ratio (MMR), which
tunes how strongly mutations (i.e., no switching of template sequences) are
preferred over recombination (i.e., switching between sequences) during
copying. Under a lower MMR value, the maximum-likelihood HMM copying path of a
focal genome more likely spans at least two ancestral genomes (i.e,. the focal
genome is a putative recombinant) than a single ancestral genome. Here, the MMR
was set to three in all the HMM runs (see ``Improvements'' for future potential
development of the HMM).

>>>>

The ``succinct tree sequence'' is an ARG data structure
that is both general (in terms of the types of ancestry that can
be described) and computationally efficient~\citep{Wong2023-efficient}.
Originally developed to facilitate large-scale coalescent
simulations~\citep{Kelleher2016-wk}, the methods have been
extended and applied to forwards-time
simulations~\citep{Kelleher2018-xc,Haller2018-tree},
calculation of population genetics statistics~\citep{Ralph2020-efficiently}
and ARG inference~\citep{Kelleher2019-ba,Wohns2022-th}.
The succinct tree sequence is based on a simple tabular representation,
which defines a set of nodes, edges, sites and mutations. A \emph{node}
represents a particular genome, which may be an observed sample
or an inferred genetic ancestor. The genetic inheritance between
a pair of nodes along a segment of genome is defined by
the \emph{edge} $(\ell, r, p, c)$, which states that
child node $c$ inherited its genome from parent node $p$
from left coordinate $\ell$ to right coordinate $r$. A \emph{site}
defines the genomic position at which some variation occurs,
and a \emph{mutation} defines the site, node and derived state
that generates variation.
Given these tables, we can then efficiently construct the local
genealogical trees along the genome (arising from recombination),
and also perform a range of calculations efficiently by
reasoning about the differences between these local
trees~\citep{Kelleher2016-wk,Ralph2020-efficiently}. These
algorithms have led to peformance increases of several orders
of magnitude over the state-of-the-art in a range of
applications~\citep{Kelleher2016-wk,Kelleher2018-xc,Kelleher2019-ba,
Ralph2020-efficiently,Baumdicker2022-ep}.
The succinct tree sequence encoding
is also very concise, allowing, for example, for millions of
complete human genomes to potentially be stored in a few gigabytes of
space~\citep{Kelleher2019-ba}.

@@

% OLD text, ready to be adapted (and deleted once done)
% For this
% application, we have parameterised the HMM using a mismatch ratio (MMR), which
% tunes how strongly mutations (i.e., no switching of template sequences) are
% preferred over recombination (i.e., switching between sequences) during
% copying. Under a lower MMR value, the maximum-likelihood HMM copying path of a
% focal genome more likely spans at least two ancestral genomes (i.e,. the focal
% genome is a putative recombinant) than a single ancestral genome. Here, the MMR
% was set to three in all the HMM runs .

In \texttt{sc2ts} we use the efficient ARG-based implementation of the
LS Viterbi algorithm from \texttt{tsinfer}~\citep{Kelleher2019-ba} to find
the most likely copying path for each sample sequence
among all sequences (sampled and inferred) in the current ARG.
In the majority of cases, with non-recombinant sample sequences,
the most likely solution is to copy from one
of the nodes in the ARG that minimised the number of mutations required
on the query sequence. Importantly, because the reference panel here consists of
\emph{every} node in the ARG, we can match to both older sample
sequences or internal nodes representing an inferred ancestral sequence
(see subsequent subsections for details about how these are added).
Thus, when no recombination is present, the LS Viterbi algorithm is
implementing a version of parsimony, in which we are guaranteed to
find a sequence that minimises the number of additional mutations required
to incorporate a newly-added sample into the ARG.
Recombination is then inferred when the most likely solution to the LS
HMM is to copy from more than one ARG node along the genome, for a
given sample sequece. [TODO MORE]

We can solve this massive optimisation problem exactly because the ARG-based
implementation of the LS HMM used in \texttt{tsinfer} scales approximately
logarithmically with reference panel size (as opposed to linearly,
for standard matrix-based approaches)~\citep{Kelleher2019-ba}.
This efficient HMM algorithm is the main reason for \texttt{tsinfer}'s
scalability, and here allows us to find closely matching
sequences and recombination paths among millions of SARS-CoV-2
genomes exactly under a well-defined statistical model.

It is important to note that the Viterbi algorithm only returns
\emph{one of} the copying paths that maximise the likelihood under
 the given mutation and recombination parameters. There may be many
such paths, from which we choose one arbitrarily. Also, the
present choice of using a single mismatch-ratio parameter to control
the likelihood of recombination vs mutation may lead to relatively
flat likelihood spaces where many different paths have equal likelihood.
There are many possibilities in using established HMM methodology
to reason about and explore the space of possible matches, which may
be a fruitful avenue for future work. See the Discussion for further
potential improvements that may be made to the HMM.
